---
title: "Getting Started with phasic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with phasic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The **phasic** package provides high-performance algorithms for computing with phase-type distributions using graph-based methods. This vignette introduces the basic functionality.

## Installation

Install the Python package first:

```bash
pip install phasic
```

Then install the R package:

```{r}
# Install from GitHub (when available)
# devtools::install_github("munch-group/phasic", subdir = ".")

# Or install locally
devtools::install_local("path/to/phasic")
```

Load the package:

```{r}
library(phasic)
```

## Creating a Simple Graph

The most basic way to create a phase-type distribution is to manually build a graph:

```{r}
# Create a graph with 1-dimensional state vectors
g <- create_graph(state_length = 1)

# Get the starting vertex
v0 <- g$starting_vertex()

# Create additional vertices
v1 <- g$find_or_create_vertex(c(1))
v2 <- g$find_or_create_vertex(c(2))

# Add edges (transitions) with rates
# Note: You would need to access the vertex methods to add edges
# This requires understanding the Python API structure
```

## Using Callbacks for Graph Construction

A more powerful approach is to use callbacks that define the transition structure:

```{r}
# Define a simple Markov chain
simple_callback <- function(state) {
  n <- state[1]
  if (n <= 0) {
    return(list())  # Absorbing state
  }
  # Transition to n-1 with rate n
  list(list(c(n - 1), n))  # 2-tuple: (next_state, rate)
}

# Create graph using callback
g <- create_graph(
  callback = simple_callback,
  state_length = 1,
  nr_samples = 5
)
```

## Coalescent Model Example

The coalescent is a common model in population genetics. Here's how to build it:

```{r}
# Coalescent callback: n lineages coalesce to n-1
coalescent_callback <- function(state) {
  n <- state[1]
  if (n <= 1) {
    return(list())  # Absorbing state (one lineage left)
  }

  # Coalescent rate: n choose 2
  rate <- n * (n - 1) / 2

  # Return: (next_state, base_weight, edge_parameters)
  # For parameterized edges, edge_parameters is coefficient vector
  list(list(c(n - 1), 0.0, c(rate)))
}

# Build parameterized coalescent graph
coal_graph <- create_graph(
  callback = coalescent_callback,
  parameterized = TRUE,
  state_length = 1,
  nr_samples = 5  # Start with 5 lineages
)

cat("Created coalescent graph with", coal_graph$vertices_length(), "vertices\n")
```

## Computing PDF

Once you have a graph, you can compute the probability density function:

```{r}
# For parameterized graphs, you need to instantiate with concrete parameters first
# We'll use trace-based approach (more efficient)
trace <- record_elimination_trace(coal_graph, param_length = 1)

# Instantiate with theta = 2.0 (scaled coalescent rate)
concrete_graph <- instantiate_from_trace(trace, theta = c(2.0))

# Compute PDF at various times
times <- seq(0.1, 5.0, length.out = 50)
pdf_values <- concrete_graph$pdf(times)

# Plot
plot(times, pdf_values, type = "l",
     xlab = "Time", ylab = "Density",
     main = "Coalescent Distribution (n=5, theta=2.0)")
```

## Computing Moments

Phase-type distributions have closed-form moment formulas:

```{r}
# Compute first 3 moments
moments <- concrete_graph$moments(power = 3)

cat("Mean:", moments[1], "\n")
cat("Second moment:", moments[2], "\n")
cat("Third moment:", moments[3], "\n")

# Variance = E[X^2] - E[X]^2
variance <- moments[2] - moments[1]^2
cat("Variance:", variance, "\n")
```

## Reward Transformation

Reward transformation allows you to model accumulated rewards instead of time:

```{r}
# Define rewards for each vertex (e.g., based on state)
rewards <- seq(1, coal_graph$vertices_length())

# Apply reward transformation
reward_graph <- concrete_graph$reward_transform(rewards)

# Compute PDF of rewarded distribution
reward_pdf <- reward_graph$pdf(times)

plot(times, reward_pdf, type = "l",
     xlab = "Accumulated Reward", ylab = "Density",
     main = "Rewarded Phase-Type Distribution")
```

## Next Steps

- See `vignette("svgd-inference")` for Bayesian inference
- See the Python API documentation for more advanced features
- Check the package documentation with `?phasic` or `help(package = "phasic")`

## References

Røikjer, T., Hobolth, A., & Munch, K. (2022). Phase-type distributions in population genetics. *Statistics and Computing*, 32(5), 1-21. https://doi.org/10.1007/s11222-022-10155-6
